Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-26 Manuscript/manuscript.Rmd
fixed methods; exchanged IR for PI

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non boundary regions. Here, the proportion of imbalance is defined as the ratio of the set of TAD boundary regions to the sum of both TAD and non-TAD boundary regions.<!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Such heavily imbalanced datasets are known to adversely impact the performance of classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

Class imbalance was characterized by each dataset's imalance ratio (IR), defined as the ratio of majority to minority sets given by the class of genomic bins:

$$IR = \dfrac{C_{majority}}{C_{minority}}$$

where $C_{majority}$ denotes the set of genomic bins that did not overlap with a flanked TAD boundary and $C_{majority}$ denotes the complementary set of genomic bins that did overlap with a flanked TAD boundary. In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling). They are described in detail below.

>>>>

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These TAD boundary regions make up the minority class, while the non-TAD boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. A dataset is considered imbalanced if PI<0.5. <!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

Class imbalance was characterized by each dataset's *proportion of imbalance* (PI), defined as the ratio of minority class observations to the sum of minority and majority class observations:

$$PI = \dfrac{C_{min}}{C_{min} + C_{maj}}$$

where $C_{min}$ denotes the complementary set of genomic bins that did overlap with a flanked TAD boundary and $C_{maj}$ denotes the set of genomic bins that did not overlap with a flanked TAD boundary. In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling). They are described in detail below.
